---
title: "xgxr issue #72: `breaks` ignored the position scale"
subtitle: "Before/after demonstration for the 1.1.6 fix"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: false
    theme: flatly
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = FALSE, message = FALSE, warning = FALSE,
  fig.width = 9, fig.height = 4.2, dpi = 110
)
library(ggplot2)
```

## The short version

`xgx_stat_ci()` lets you pass `breaks` to control how the x axis is binned.
If you also put the x axis on a log scale with `xgx_scale_x_log10()`, the
binning was wrong: most of your data was silently dropped and the remaining
points were placed in the wrong bins.

**This was not a ggplot2 4.0 regression**, despite what the issue title says.
It reproduces identically under ggplot2 3.5.1. Anyone combining `breaks` with a
log x axis has been getting quietly incorrect plots for years.

A second, separate bug affected `distribution = "ordinal"` **even with no log
scale at all** — the smallest observation in the dataset fell outside every bin
and was drawn as a spurious extra point.

## How this document is built

Nothing here is a mock-up or a hand-written approximation of the old behaviour.
This document checks the last pre-fix commit out into a temporary git worktree,
runs the identical analysis against both that version and the current working
tree in separate R subprocesses, and plots what each one actually returns.

That means it needs to be knit from inside a clone of the xgxr repository, and
it will keep working as `xgx_stat_ci` continues to evolve.

```{r run-probe}
BEFORE_REF <- "2f8f6872142c159071308b60cb5794834336c406"

repo <- normalizePath(
  system2("git", c("rev-parse", "--show-toplevel"), stdout = TRUE, stderr = FALSE)
)

# The probe. Written to a file and run once per package version, because two
# versions of the same package cannot live in one R session. It deliberately
# uses only long-standing public API so it runs against old and new code alike.
probe_src <- '
suppressMessages(library(ggplot2))
args    <- commandArgs(trailingOnly = TRUE)
pkg_dir <- args[[1]]
out_rds <- args[[2]]
suppressMessages(devtools::load_all(pkg_dir, quiet = TRUE))

set.seed(1234)
df <- data.frame(
  x = c(runif(300, 0.4, 2.8), runif(142, 2.8, 5.2),
        runif(25, 5.2, 7.6), runif(3, 7.6, 10)),
  y = rbinom(470, 1, 0.3))
brk <- quantile(df$x)

set.seed(5678)
ord <- data.frame(
  x = 10^runif(400, -0.5, 1),
  response = factor(sample(c("Mild", "Moderate", "Severe"), 400, TRUE)))
ord_brk <- quantile(ord$x)

grab <- function(p, logged) {
  d <- suppressMessages(suppressWarnings(
    tryCatch(layer_data(p), error = function(e) NULL)))
  if (is.null(d)) return(NULL)
  # On a log10 axis the stat works in transformed space, so map x back to the
  # original units to make the panels directly comparable.
  if (logged) {
    for (nm in c("x", "xmin", "xmax")) if (nm %in% names(d)) d[[nm]] <- 10^d[[nm]]
  }
  d
}

base <- ggplot(df, aes(x = x, y = y)) +
  xgx_stat_ci(breaks = brk, geom = "point")
ord_base <- ggplot(ord, aes(x = x, response = response, colour = response)) +
  xgx_stat_ci(distribution = "ordinal", breaks = ord_brk, geom = "point")

saveRDS(list(
  ggplot2    = as.character(packageVersion("ggplot2")),
  xgxr       = as.character(packageVersion("xgxr")),
  brk        = brk,
  ord_brk    = ord_brk,
  cont_lin   = grab(base, FALSE),
  cont_log   = grab(base + xgx_scale_x_log10(), TRUE),
  ord_lin    = grab(ord_base, FALSE),
  ord_log    = grab(ord_base + xgx_scale_x_log10(), TRUE)
), out_rds)
'

tmp <- file.path(tempdir(), paste0("issue72-report-", Sys.getpid()))
dir.create(tmp, showWarnings = FALSE, recursive = TRUE)
probe_file <- file.path(tmp, "probe.R")
writeLines(probe_src, probe_file)
worktree <- file.path(tmp, "before")

run_probe <- function(pkg_dir, label) {
  out_rds <- file.path(tmp, paste0(label, ".rds"))
  st <- system2("Rscript", c(shQuote(probe_file), shQuote(pkg_dir), shQuote(out_rds)),
                stdout = FALSE, stderr = FALSE)
  if (st != 0 || !file.exists(out_rds)) {
    stop("probe failed for '", label, "'. Knit this from inside the xgxr repo, ",
         "with devtools and ggplot2 installed.")
  }
  readRDS(out_rds)
}

st <- system2("git", c("-C", shQuote(repo), "worktree", "add", "--detach",
                       shQuote(worktree), BEFORE_REF),
              stdout = FALSE, stderr = FALSE)
if (st != 0) stop("could not create the git worktree for ", BEFORE_REF)

before <- run_probe(worktree, "before")
after  <- run_probe(repo,     "after")

system2("git", c("-C", shQuote(repo), "worktree", "remove", "--force",
                 shQuote(worktree)), stdout = FALSE, stderr = FALSE)

tag <- function(d, version, panel) {
  if (is.null(d) || !nrow(d)) return(NULL)
  d$version <- version
  d$panel   <- panel
  d
}
bind <- function(...) do.call(rbind, Filter(Negate(is.null), list(...)))

pal <- c("before (1.1.5)" = "#C0392B", "after (1.1.6)" = "#1B7F79")
```

The comparison below is `r substr(BEFORE_REF, 1, 10)` (**before**) against the
current working tree (**after**), both running ggplot2 `r after$ggplot2`.

---

## Bug 1 — continuous response on a log10 x axis

The dataset is 470 binary observations with x spanning 0.4 to 10. `breaks` is
set to `quantile(x)`, so there are four bins, and each should hold roughly a
quarter of the data.

The grey band is the correct answer: the same call with **no log scale**, which
was never affected by the bug.

```{r cont-plot, fig.height = 4.2}
truth <- after$cont_lin

cont <- bind(
  tag(before$cont_log, "before (1.1.5)", "with xgx_scale_x_log10()"),
  tag(after$cont_log,  "after (1.1.6)",  "with xgx_scale_x_log10()")
)

ggplot(cont, aes(x = x, y = y, colour = version)) +
  geom_ribbon(
    data = truth, aes(x = x, ymin = ymin, ymax = ymax),
    inherit.aes = FALSE, fill = "grey70", alpha = 0.30
  ) +
  geom_line(
    data = truth, aes(x = x, y = y),
    inherit.aes = FALSE, colour = "grey40", linetype = "22"
  ) +
  geom_pointrange(aes(ymin = ymin, ymax = ymax), size = 0.5) +
  facet_wrap(~version) +
  scale_x_log10() +
  scale_colour_manual(values = pal, guide = "none") +
  labs(
    x = "x (original units, log10 axis)", y = "estimated probability",
    title = "Continuous response, breaks = quantile(x)",
    subtitle = "Grey = the correct result (same call without a log scale)"
  ) +
  theme_bw() +
  theme(strip.text = element_text(face = "bold"))
```

```{r cont-table}
knitr::kable(
  data.frame(
    version  = c("before (1.1.5)", "after (1.1.6)"),
    bins     = c(nrow(before$cont_log), nrow(after$cont_log)),
    expected = length(after$brk) - 1L,
    estimates = c(
      paste(sprintf("%.4f", before$cont_log$y), collapse = ", "),
      paste(sprintf("%.4f", after$cont_log$y),  collapse = ", ")
    )
  ),
  caption = "Binned estimates with xgx_scale_x_log10()"
)
```

Before the fix two of the four bins vanish entirely and the two survivors sit at
the wrong heights. After the fix all four bins are present and the estimates
match the no-log-scale control exactly.

### Why it happened

By the time data reaches the stat, the position scale has already log10
transformed it — x values of 0.4 to 10 have become -0.4 to 1. But `breaks` is
still in the user's original units. So `cut()` compares transformed data against
untransformed break points, and nearly every observation falls outside every bin.

ggplot2's own `StatBin` has handled this for years:

```r
if (!scales[[x]]$is_discrete()) {
  breaks <- scales[[x]]$transform(breaks)
}
```

`xgxr::StatSummaryBinQuant` is a hand-copied fork of ggplot2's
`stat-summary-bin.R` — its roxygen block pins it to a 2020 ggplot2 commit — and
it simply never picked that step up. The copy contained a literal `breaks <- breaks`
where ggplot2 does the transform.

---

## Bug 2 — ordinal response, no log scale required

This one is independent and arguably more likely to have bitten someone quietly,
because it needs no log scale at all. With `distribution = "ordinal"` and three
response categories over four bins there should be exactly 12 points.

```{r ord-plot, fig.height = 4.4}
ord <- bind(
  tag(before$ord_lin, "before (1.1.5)", "x"),
  tag(after$ord_lin,  "after (1.1.6)",  "x")
)
if (!"response" %in% names(ord)) ord$response <- ord$group
ord$response <- factor(ord$response)

ggplot(ord, aes(x = x, y = y, colour = response)) +
  geom_pointrange(aes(ymin = ymin, ymax = ymax),
                  position = position_dodge(width = 0.25), size = 0.45) +
  facet_wrap(~version) +
  labs(
    x = "x (linear axis, no log scale)", y = "estimated probability",
    title = "Ordinal response, breaks = quantile(x), no log scale",
    subtitle = "The stray point at y = 1.0 on the left is the bug"
  ) +
  theme_bw() +
  theme(strip.text = element_text(face = "bold"))
```

```{r ord-table}
knitr::kable(
  data.frame(
    version = c("before (1.1.5)", "after (1.1.6)"),
    `no log scale` = c(nrow(before$ord_lin), nrow(after$ord_lin)),
    `with log10`   = c(nrow(before$ord_log), nrow(after$ord_log)),
    expected       = (length(after$ord_brk) - 1L) * 3L,
    check.names = FALSE
  ),
  caption = "Number of points returned by the ordinal stat"
)
```

Before the fix the ordinal path produced **more** rows than it should have on a
linear axis: the single smallest observation in the dataset fell into an `NA`
bin and was drawn as an extra point pinned at y = 1.0. The cause was a `cut()`
call missing `include.lowest = TRUE`. On a log axis the same path lost rows to
the transform bug as well.

The ordinal fix was structurally harder than the continuous one. Binning used to
happen in `setup_data()`, which ggplot2 never hands the scales to, so it had to
move to `compute_layer()` where `layout$panel_scales_x` is actually available.

---

## The invariant

The cleanest way to state what was broken: **adding a log10 scale to a plot must
not change the binned estimates.** It changes how the data is drawn, never what
is computed.

```{r invariant}
holds <- function(v) {
  if (is.null(v$cont_lin) || is.null(v$cont_log)) return(FALSE)
  isTRUE(all.equal(v$cont_log$y, v$cont_lin$y))
}
knitr::kable(
  data.frame(
    version   = c("before (1.1.5)", "after (1.1.6)"),
    invariant = c(if (holds(before)) "HOLDS" else "VIOLATED",
                  if (holds(after))  "HOLDS" else "VIOLATED")
  ),
  caption = "log10 scale must not change the binned estimates"
)
```

This is now asserted in `tests/testthat/test-xgx_stat_ci.R`. The tests fail
against the pre-fix code and pass after.

---

## What this means for users

- **Affected:** anyone passing `breaks` to `xgx_stat_ci()` together with
  `xgx_scale_x_log10()` (or any transformed position scale), and anyone using
  `distribution = "ordinal"` with `breaks` on any scale.
- **Not affected:** the default binning via `bins`, and any plot without a
  transformed x scale.
- **Workaround on 1.1.5 and earlier:** `coord_transform(x = "log10")` instead of
  the scale. It doesn't transform the data before the stat sees it, so the bug
  never triggers. It still works after the fix, where the new code is simply a
  no-op.

One deliberate behaviour change: on a log10 axis each point now sits at the
**geometric** midpoint of its bin rather than the arithmetic one. That is
`bin_loc()` operating in transformed space, and it means the point appears at
the visual centre of its bin on the axis as displayed — which is what you want,
but it will shift point positions slightly in existing plots.

## The wider issue

The root cause is the copied-fork pattern. `StatSummaryBinQuant`,
`StatSummaryOrdinal` and `StatSmoothOrdinal` are all snapshots of ggplot2
internals taken at various points, and the package also reaches into
`ggplot2:::bin_loc`, `dapply`, `make_summary_fun`, `snake_class` and
`check_required_aesthetics`. Every one of those is a place where ggplot2 can fix
or change something and xgxr silently keeps the old behaviour. This bug is
exactly that failure mode. Worth a conversation about how much of it we can
retire.

---

*Fix is in [PR #73](https://github.com/Novartis/xgxr/pull/73), released in xgxr 1.1.6.
Issue: [Novartis/xgxr#72](https://github.com/Novartis/xgxr/issues/72).*
