---
title: "xgxr"
author: "Matt Fidler"
date: "11/12/2022"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(testthat)
```

# Brief Package Description


Supports a structured approach for exploring PKPD data <https://opensource.nibr.com/xgx/>. It also contains helper functions for enabling the modeler to follow best R practices (by appending the program name, figure name location, and draft status to each plot). In addition, it enables the modeler to follow best graphical practices (by providing a theme that reduces chart ink, and by providing time-scale, log-scale, and reverse-log-transform-scale functions for more readable axes). Finally, it provides some data checking and summarizing functions for rapidly exploring pharmacokinetics and pharmacodynamics (PKPD) datasets.

# Datasets used for testing

This describes the datasets (if any) or functions (if any) that are used
to test the package.

`nlmixr2data::theo_sd`


# Functions that are exported

## GGproto functions

These functions are exported but are not expected to be used directly by the user and are not tested:

- `StatSmoothOrdinal`
- `StatSummaryBinQuant`
- `StatSummaryOrdinal`

## Interactive only functions

These functions are based on interactive user input and are not tested:

- `edit_rmd_template_str`

## Exported from other packages

- `gtable`
- `nlsLM`
- `predict.nls`

## Functions that are display only, and do no calculation

- `theme_xgr` -- changes the ggplot2 theme to black and white

- `xgx_annotate_filenames` -- adds footnotes to figuers

- `xgx_annotate_status` -- Adds "DRAFT" to figures

- `xgx_annotate_status_png` -- nadds status to the png files

- `xgx_theme`

- `xgx_theme_set`

# Testible functions without graphics

- `xgx_breaks_log10` display log-10 breaks on scale

```{r}
test_that("log10 breaks make sense", {
  expect_equal(xgx_breaks_log10(c(0.001, 2000)), 
               c(0.001, 0.01, 0.1, 1, 10, 100, 1000))
})

```

- `xgx_breaks_time` displays breaks with time

```{r}
test_that("xgx_breaks_time makes sense", {
  expect_equal(xgx_breaks_time(c(0, 24*10), "h"),
               c(0, 60, 120, 180, 240))
               
})

```
- `xgx_check_data`

```{r}
test_that("xgx_check_data produces no errors", {
  expect_error(xgx_check_data(nlmixr2data::theo_sd), NA)
})
```

- `xgx_conf_int`

```{r}
test_that("xgx_conf_int", {
  withr::with_seed(24,{
    expect_equal(xgx_conf_int(rnorm(1000)),
                 structure(list(y = -0.0469587129006418, ymin = -0.109521192672548, ymax = 0.0156037668712641), class = "data.frame", row.names = c(NA, -1L)))
  })
})

```   

- `xgx_dirs2char`

```{r}

test_that("xgx_dirs2char", {
  dirs <- list(parent_dir = "/your/parent/path/",
            rscript_dir = "./Rscripts/",
            rscript_name = "Example.R",
            results_dir = "./Results/",
            filename = "your_file_name.png")
caption <- xgx_dirs2char(dirs)
expect_equal("/your/parent/path/\n./Rscripts//Example.R\n./Results//your_file_name.png\nCreated: 2022-11-07 20:10:34",
             caption)
})

```

- `xgx_labels_log10 `

```{r}
test_that("xgx_breaks_time makes sense", {
  expect_equal(xgx_labels_log10(xgx_breaks_log10(c(0.001, 2000))),
               c("0.001", "0.01", "0.1", "1", "10", "100", "1000"))
})
```

- `xgx_minor_breaks_log10`

```{r}

test_that("xgx_breaks_time makes sense", {
  expect_equal(xgx_minor_breaks_log10(c(0.001, 2000)),
               c(0.002, 0.003, 0.004, 0.005, 0.006, 0.007, 0.008, 0.009, 
                 0.02, 0.03, 0.04, 0.05, 0.06, 0.07, 0.08, 0.09, 
                 0.2, 0.3, 0.4, 0.5, 0.6, 0.7, 0.8, 0.9, 
                 2, 3, 4, 5, 6, 7, 8, 9, 
                 20, 30, 40, 50, 60, 70, 80, 90, 
                 200, 300, 400, 500, 600, 700, 800, 900, 
                 2000))
})
```

# Interfaces that allow creation of ggplots

- `xgx_plot` -- This simply is a wrapper around `ggplot`, not sure the rationale.  This is a display-only style function


# Save ggplot2 objects

- `xgx_save` -- Save ggplot

- `xgx_save_table` -- saves table as image

# Scales/stats/geom

These produce graphical output and need to be tested further in the future, probably using `vdiffr`

- `xgx_geom_ci`
- `xgx_geom_pi`
- `xgx_geom_smooth`, use "polr", "nls", "nlsLM"
- `xgx_geom_smooth` uses `minpack.lm::nlsLM`
- `xgx_scale_x_log10`
- `xgx_scale_x_percentchangelog10`
- `xgx_scale_x_reverselog10`
- `xgx_scale_x_time_units`
- `xgx_scale_y_log10`
- `xgx_scale_y_percentchangelog10`
- `xgx_scale_y_reverselog10`
- `xgx_scale_y_time_units`
- `xgx_stat_ci`
- `xgx_stat_pi`
- `xgx_stat_smooth`
- `xgx_summarize_covariates`
- `xgx_summarize_data`
